---
title: "Geography 13"
author: "[Anisah Calvin Cuevas](https://acuevas917.github.io/)"
subtitle: 'Lab 03: Distances and the Border Zone'
output:
  html_document:
    theme: journal
---

```{r, message=F, warning=F}
# SPDS
library(tidyverse)
library(sf)
library(units)
library(dplyr)
library(readr)
library(knitr)

# Data
library(USAboundaries)
library(rnaturalearth)

# Visualization
library(ggplot2)
library(gghighlight)
library(ggrepel)
library(knitr)
```

### Question 1

#### 1.1 Define a Projection
```{r, message=FALSE, warning=FALSE}
eqdc = '+proj=eqdc +lat_0=40 +lon_0=-96 +lat_1=20 +lat_2=60 +x_0=0 +y_0=0 +datum=NAD83 +units=m +no_defs'
```

#### 1.2 USA State Boundaries
```{r, message=FALSE, warning=FALSE}
USAboundaries::us_states(resolution = "low") %>%
  st_as_sf(coords = c("lng", "lat"), crs = 4326) %>%
  filter(!state_name %in% c("Puerto Rico",
                            "Alaska",
                            "Hawaii")) ->
  conus
st_transform(conus, eqdc) -> conus
```

#### 1.3 Mexico, United States, and Canada
```{r, message=FALSE}
rnaturalearthdata::countries110 -> countries110
st_as_sf(countries110, crs = 4326) -> worldboundaries
st_transform(worldboundaries, eqdc) -> worldboundaries
```

#### 1.4 City Locations
```{r, message=FALSE}
readr::read_csv("~/github/geog-13-labs/data/US Cities.csv") %>%
  st_as_sf(coords = c("lng", "lat"), crs = 4326) %>%
  filter(!state_name %in% c("Hawaii", "Alaska", "Puerto Rico")) %>%
         st_transform(., eqdc) ->
  uscities
```

### Question 2

#### 2.1 Distance to USA Border
```{r}
st_union(conus) %>%
  st_cast("MULTILINESTRING") ->
  USboundaries
uscities %>%
  select(city, state_name) %>%
  mutate(dist = st_distance(uscities, USboundaries),
  dist = units::set_units(dist, "km"),
  dist = units::drop_units(dist)) ->
  border_dist
border_dist %>%
  slice_max(dist, n = 5) %>%
  st_drop_geometry() ->
  top5_to_border
knitr::kable(top5_to_border,
             caption = "Distance of US Cities to US Borders",
             col.names = c("City", "State", "Distance to Border"))
```

#### 2.2 Distance to States
```{r}
st_combine(conus) %>%
  st_cast("MULTILINESTRING") ->
  USboundaries2
uscities %>%
  select(city, state_name) %>%
  mutate(dist2 = st_distance(uscities, USboundaries2),
         dist2 = units::set_units(dist2, "km"),
         dist2 = units::drop_units(dist2)) ->
  border_dist2
border_dist2 %>%
  slice_max(dist2, n = 5) %>%
  st_drop_geometry() ->
  top5_to_stateborder
knitr::kable(top5_to_stateborder,
             caption = "Distance of US Cities to US State Borders",
             col.names = c("City", "State", "Distance to State Border"))
```

#### 2.3 Distance to Mexico
```{r}
worldboundaries %>%
  st_cast("MULTILINESTRING") %>%
  filter(admin == "Mexico") ->
  mexico_border
uscities %>%
  select(city, state_name) %>%
  mutate(dist3 = st_distance(uscities, mexico_border),
         dist3 = units::set_units(dist3, "km"),
         dist3 = units::drop_units(dist3)) ->
  border_dist3
border_dist3 %>%
  slice_max(dist3, n = 5) %>%
  st_drop_geometry() ->
  top5_to_mexico
knitr::kable(top5_to_mexico,
             caption = "Distance of US Cities to Mexican Border",
             col.names = c("City", "State", "Distance to Mexican Border"))
```

#### 2.4 Distance to Canada
```{r}
worldboundaries %>%
  st_cast("MULTILINESTRING") %>%
  filter(admin == "Canada") ->
  canada_border
uscities %>%
  select(city, state_name) %>%
  mutate(dist4 = st_distance(uscities, canada_border),
         dist4 = units::set_units(dist4, "km"),
         dist4 = units::drop_units(dist4)) ->
  border_dist4
border_dist4 %>%
  slice_max(dist4, n = 5) %>%
  st_drop_geometry() ->
  top5_to_canada
knitr::kable(top5_to_canada,
             caption = "Distance of US Cities to Canadian Border",
             col.names = c("City", "State", "Distance to Canadian Border"))
```

### Question 3

#### 3.1 Data
```{r, warning=FALSE}
uscities %>%
  slice_max(population, n = 10) ->
  top10_pop
ggplot() +
  geom_sf(data = conus) +
  geom_sf(data = top10_pop$geometry, add = TRUE, size= 2, color = "red") +
  ggrepel::geom_label_repel(
    data = top10_pop,
    aes(label = city, geometry = geometry),
    stat = "sf_coordinates",
    size = 3) +
  labs(title = "10 Largest US Cities by Population") +
  ggthemes::theme_map() 
```

#### 3.2 City Distance from Border
```{r}
border_dist %>%
  slice_max(dist, n = 5) ->
  top5_to_bordermap
ggplot() +
  geom_sf(data = USboundaries) +
  geom_sf(data = border_dist, aes(col = dist), size = .1) +
  geom_sf(data = top5_to_bordermap, col = "red", size = 1) +
  scale_color_gradient(low = "grey", high = "blue") +
  ggrepel::geom_label_repel(
    data = top5_to_bordermap,
    aes(label = city, geometry = geometry),
    stat = "sf_coordinates",
    size = 3) +
  labs(title = "US Cities' Distance from National Border") +
  ggthemes::theme_map()
```

#### 3.3 City Distance from Nearest State
```{r}
border_dist2 %>%
  slice_max(dist2, n = 5) ->
  top5_to_statebordermap
ggplot() +
  geom_sf(data = USboundaries2) +
  geom_sf(data = border_dist2, aes(col = dist2), size = .1) +
  geom_sf(data = top5_to_statebordermap, col = "red", size = 1) +
  scale_color_gradient(low = "grey", high = "blue") +
  ggrepel::geom_label_repel(
    data = top5_to_statebordermap,
    aes(label = city, geometry = geometry),
    stat = "sf_coordinates",
    size = 3) + 
  labs(title = "US Cities' Distance from State Borders") +
  ggthemes::theme_map()
```

#### 3.4 Equidistance Boundary from Mexico and Canada
```{r, warning=FALSE}
uscities %>%
  select(city, state_name, population) %>%
  mutate(dist3 = st_distance(uscities, mexico_border),
         dist3 = units::set_units(dist3, "km"),
         dist3 = units::drop_units(dist3)) %>%
  mutate(dist4 = st_distance(uscities, canada_border),
         dist4 = units::set_units(dist4, "km"),
         dist4 = units::drop_units(dist4)) %>%
  mutate(diffdist = abs(dist3 - dist4)) ->
  equaldistance
equaldistance %>% 
  filter(diffdist <= 100) %>%
  slice_max(population, n = 5) ->
  ED_pop
ggplot() +
  geom_sf(data = USboundaries2) +
  geom_sf(data = equaldistance, aes(col = diffdist), size = .1) +
  scale_color_gradient(low = "grey", high = "blue") +
  gghighlight::gghighlight(diffdist <= 100) +
  ggrepel::geom_label_repel(
    data = ED_pop,
    aes(label = city, geometry = geometry),
    stat = "sf_coordinates",
    size = 3) +
  labs(title = "Cities that are equal distance from the Canadian and Mexican border ± 100 km") +
  ggthemes::theme_map()
```

### Question 4

#### 4.1 Quantifing Border Zone
```{r}
uscities %>%
  select(city, state_name, population) %>%
  mutate(dist = st_distance(uscities, USboundaries),
         dist = units::set_units(dist, "km"),
         dist = units::drop_units(dist)) %>%
  filter(dist <= 160) %>%
  st_drop_geometry() ->
  bordercities
bordercities %>%
  summarize(population) %>%
  sum()
data.frame(number_cities = "12283",
           pop = "259935815") -> table
knitr::kable(table,
             caption = "Cities Near National Border",
             col.names = c("Number of Cities", "Population"))
```

#### 4.2 Mapping Border Zone
```{r}
ggplot() + 
  geom_sf(data = conus, aes(fill = aland), lty = 3, alpha = .5) + 
  scale_fill_gradient(low = 'blue', high = "red") +
  gghighlight(aland > 5e10) +
  geom_sf(data = conus, size = 1, fill = "NA") +
  geom_sf(data = top10_pop, size= 2, color = "red") +
  ggrepel::geom_label_repel(
    data = top10_pop,
    aes(label = city, geometry = geometry),
    stat = "sf_coordinates",
    size = 3) +
  ggthemes::theme_map() 
```
