---
title: "Geography 13"
author: "[Anisah Calvin Cuevas](https://acuevas917.github.io/)"
subtitle: 'Lab 02: COVID-19 Pandemic'
output:
  html_document:
    theme: journal
---
### Question 1
```{r, message=F, warning=F}
library(tidyverse)
library(knitr)
library(readxl)
library(zoo)
library(dplyr)

covid19 = read_csv('https://raw.githubusercontent.com/nytimes/covid-19-data/master/us-counties.csv')
```

#### California Data
```{r, message=F, warning=F}
covid_ca = covid19 %>%
  filter(state=="California") %>%
  group_by(county) %>%
  mutate(newcases = cases - lag(cases)) %>%
  ungroup()
```

#### Top 5 Cumulative
```{r, message=F, warning=F}
top5cumulative= covid_ca %>%
  filter(date==max(date)) %>%
  slice_max(cases, n=5) %>%
  select(county, cases)
knitr::kable(top5cumulative,
             caption = "Most Cumulative Cases California Counties",
             col.names = c("County", "Cumulative Cases"))
```

#### Top 5 with New Cases
```{r, message=F, warning=F}
top5_newcases = covid_ca %>%
  filter(date==max(date)) %>%
  slice_max(newcases, n=5) %>%
  select(county, newcases)
knitr::kable(top5_newcases,
             caption = "Most New Cases California Counties",
             col.names = c("County", "New Cases"))
```

```{r, message=F, warning=F}
pop_est = read_excel("~/github/geog-13-labs/data/PopulationEstimates.xls",
                     skip=2)
pop_est = pop_est %>%
  select_(fips = "FIPStxt", state = "State", "Area_Name", pop2019 = "POP_ESTIMATE_2019")
pop_joined_covid = inner_join(pop_est, covid_ca, by = "fips")
cases_percapita = pop_joined_covid %>%
  filter(date==max(date)-13) %>%
  mutate(most_percapita = (sum(cases))/pop2019) %>%
  mutate(new_percapita = ((sum(cases-lag(cases)))/pop2019))
```

#### Most Cases Per Capita
```{r, message=F, warning=F}
most_cumulative_percapita = cases_percapita %>%
  slice_max(most_percapita, n=5) %>%
  select(county, most_percapita)
knitr::kable(most_cumulative_percapita,
             caption = "Most Cumulative Cases Per Capita California Counties",
             col.names = c("County", "Cumulative Cases per Capita"))
```

#### Most New Cases Per 100,000 People
```{r, message=F, warning=F}
last14days = pop_joined_covid %>%
  filter(date>max(date)-14) %>%
  group_by(county, pop2019) %>%
  summarise(newcases=sum(newcases)) %>%
  ungroup() %>%
  mutate(case_per100k = newcases/(pop2019/100000)) %>%
  filter(case_per100k<=100)
knitr::kable(last14days,
             caption="Counties with Most New Cases")
```

#### Analysis
1. Counties with greater area have the most cumulative cases due to a large population
2. Counties with greater area have the most new cases due to a large population
3. Counties with lesser area have for cases per capita/100,000 people due to a small population

### Question 2
#### Rolling Mean Graph
```{r}
covid19 %>%
  group_by(state, date) %>%
  summarize(cases = sum(cases)) %>%
  ungroup() %>%
  filter(state %in% c("New York", "California", "Louisiana", "Florida")) %>%
  group_by(state) %>%
  mutate(newCases = cases - lag(cases)) %>%
  mutate(avg = zoo::rollmean(newCases, 7, fill = NA, allign = "right")) ->
  covid_step1 
covid_step1 %>%
  ggplot(aes(x = date, group = 1)) + 
  geom_col(aes(y = newCases), col = NA, fill = "#F5B8B5") +
  geom_line(aes(y = avg), col = "blue", size = 1) + 
  facet_wrap(~state) + 
  ggthemes::geom_rangeframe() +
  labs(title = "Seven Days Average of New Cases",
       x = "Date", y = "New Cases") + 
  theme(aspect.ratio = .7)
```

```{r}
pop_est %>%
  select_(fips = "fips", state = "state", "Area_Name", pop2019 = "pop2019") %>%
  mutate(fips = as.integer(fips)) %>%
  filter(state %in% c("NY","CA","LA", "FL")) ->
  new_pop4
```


```{r}
covid19 %>%
  filter(state %in% c("New York", "California", "Louisiana", "Florida")) %>%
  group_by(state)%>%
  mutate(fips = as.integer(fips)) %>%
  mutate(newCases = cases - lag(cases)) %>%
  ungroup() ->
  new_covid4
```

```{r}
right_join(new_covid4, new_pop4, by = "fips") -> covidPop4
```

```{r}
covidPop4 %>%
  group_by(state.x, date) %>%
  mutate(newCasesPerCapita4 = newCases/pop2019) %>%
   mutate(avg4 = zoo::rollmean(newCasesPerCapita4, 7, fill = NA, allign = "right"))
```

#### Analysis
This graph was scaled by population, and it can be seen that Louisiana's situation seems less dire to California due to differences in population between the states. 
